---
title: "Using DESeq2 in R"
author: Daniel Neves and Daniel Sobral
date: "October 10, 2018"
output: 
  html_document: 
    keep_md: yes
---

```{r, eval=FALSE, echo=FALSE}
knitr::purl("tutorial1.Rmd")
```

This document demonstrates how to use *DESeq2* in the *R environment* to perform a differential expression analysis using the the Trapnell datasets as an example. We will first need to tell R what samples are going to be analysed, then run the *DESeq2* pipeline and plot the results of the analysis.

# Setting up the environment

First we need to make sure that R is running on the same directory where we placed the counts files (the files called trapnell_counts_C1_R1.tab, trapnell_counts_C1_R2.tab, etc...). To do this either type `setwd("path/to/directory")` in the R console, or use the *Files* panel to navigate to the counts directory and then select *More -> Set As Working Directory*.

# Setting up the count data and metadata

In this example, instead of loading the sample counts ourselves, we are going to let *DESeq2* handle that for us. For this, we just need to tell *DESeq2* what files correspond to each sample. We start by setting variables to hold the list of samples we are going to analyze. We create a list of sample names, a list of sample files (where the counts are), and a list of experimental conditions, telling which samples correspond to each condition. Type the following lines in the R console and press Enter.

```{r}
sampleNames <- c("trapnell_counts_C1_R1", "trapnell_counts_C1_R2", "trapnell_counts_C1_R3", "trapnell_counts_C2_R1", "trapnell_counts_C2_R2", "trapnell_counts_C2_R3")

sampleFiles <- c("trapnell_counts_C1_R1.tab", "trapnell_counts_C1_R2.tab", "trapnell_counts_C1_R3.tab", "trapnell_counts_C2_R1.tab", "trapnell_counts_C2_R2.tab", "trapnell_counts_C2_R3.tab")

sampleConditions <- c("C1", "C1", "C1", "C2", "C2", "C2")
```

We can confirm the values in these variables by simply typing a variable name in the R console and pressing Enter.

```{r}
sampleNames
sampleFiles
sampleConditions
```

For convenience, we place this information in a table variable that we call `sampleTable`.

```{r}
sampleTable <- data.frame(sampleName = sampleNames,
                          fileName = sampleFiles,
                          condition = sampleConditions)

sampleTable
```

# Running a differential expression test with DESeq2

With the sample table prepared, we are ready to run **DESeq2**. First need to import it into the R environment. This is done with the `library` command.

```{r, message=FALSE}
library("DESeq2")
```

Then, we prepare a special structure to tell *DESeq2* what samples we are going to analyse (our sample table), and what comparison we are goind to make. By setting the `design`argument to `~ condition`, we are specifying that column as the experimental variable (C1 or C2).

```{r}
ddsHTSeq <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable,
                                       design= ~ condition)
```

We can run the whole DESeq2 pipeline with a single command. This will perform normalization of the raw counts, estimate variances, and perform the differential expression tests.

```{r}
ddsHTSeq <- DESeq(ddsHTSeq)
```

We can then extract the results of the differential expression in the form of a table using the `results` function. The `head` function will print the first lines of this table on the console.

```{r}
resHTSeq <- results(ddsHTSeq)

head(resHTSeq)
```

*Hint: you can type `View(resHTSeq)` to open the full table in a separate window*

We can ask how many genes are differentially expressed (using a cutoff of 0.05) with this command.

```{r}
table(resHTSeq$padj < 0.05)
```

**Question**: How many genes have padj less than 0.01? How many genes have nominal p-values less than 0.01?

<details><summary><b>Click Here to see the answer</b></summary>

```{r}
table(resHTSeq$padj < 0.01)
table(resHTSeq$pvalue < 0.01)
```

</details>

---

We sort this table by p-value (smaller p-values on top), and save it to a file so that we can later import it into Excel. 

```{r}
orderedRes <- resHTSeq[ order(resHTSeq$padj), ]

write.csv(as.data.frame(orderedRes), file="trapnell_C1_VS_C2.DESeq2.csv")
```

We can also retrieve and save a table of normalized counts. 

```{r}
normCounts <- counts(ddsHTSeq, normalized = TRUE)

head(normCounts)

write.csv(as.data.frame(orderedRes), file="trapnell_normCounts.DESeq2.csv")
```

Finally, we can merge the two tables into a single one.

```{r}
merged.results <- merge(normCounts, orderedRes, by="row.names")

head(merged.results)
```

**Exercise**: When merging the two tables, we lost the ordering by p-value. Can you reorder the `merged.results` table by p.value?

<details><summary><b>Click Here to see the answer</b></summary>

```{r}
merged.results <- merged.results[ order(merged.results$padj), ]
head(merged.results)
```

</details>

# Visualizing results

*DESeq2* provides several functions to visualize the results, while additional plots can be made using the extensive R graphics capabilities. Visualization can help to better understand the results, and catch potential problems in the data and analysis. We start here by reproducing the plots that we previously obtained using Galaxy.

## Dispersion plot

We can plot the *DESeq2* dispersion re-estimation procedure by typing:

```{r}
plotDispEsts(ddsHTSeq)
```

## P-value distribution

As a sanity check, we can inspect the distribution of p-values using the `hist` function.

```{r}
hist(resHTSeq$pvalue, breaks=0:50/50, xlab="p value", main="Histogram of nominal p values")
```

## MA-plot

To make an (unshrunken) **MA-plot**, that displays the relationship between a genes' mean expression and its fold-change between experimental conditions, type the following in the R console. 

```{r}
plotMA(resHTSeq)
```

To obtain an **MA-plot** with shrunken log2 fold-changes we use the `lfcShrink` function. This function is equivalent to the `results` function that we called previously, but will return a table with the *log2FoldChange* and *lfcSE* columns replaced with the shrunken values. The `coef` argument is used to specify what *contrast* we are interested in analysing (in this case condition_C2_vs_C1), so we first call `resultsNames` to determine the right coefficient.

```{r}
resultsNames(ddsHTSeq)
resHTSeqShrunk <- lfcShrink(ddsHTSeq, coef=2)
plotMA(resHTSeqShrunk)
```

## Volcano plot

A **Volcano plot** displays the relationship between fold-change and evidence of differential expression (represented as -log p-adusted). *DESeq2* doesn't provide a function to display a **Volcano plot**, but we can create one using R's base plot functions. In red we highlight genes differentially expressed with Padj < 0.05.

```{r}
highlight <- which(resHTSeq$padj < 0.05)

plot(resHTSeq$log2FoldChange, -log10(resHTSeq$pvalue), xlab="log2 Fold-change", ylab="-log P-adjusted", pch=20, cex=0.5)
points(resHTSeq$log2FoldChange[ highlight ], -log10(resHTSeq$pvalue[ highlight ]), col="red", pch=20, cex=0.5)
abline(v=0, h=-log10(0.05), lty="dashed", col="grey")
```

**Exercise**: Change the commands above to make a **volcano plot** using the shrunken log fold changes instead. Also change the threshold of differential expression to 0.01 and the color of the differentially expressed genes to green.

<details><summary><b>Click Here to see the answer</b></summary>

```{r}
highlight <- which(resHTSeqShrunk$padj < 0.01)

plot(resHTSeqShrunk$log2FoldChange, -log10(resHTSeqShrunk$pvalue), xlab="shrunken log2 Fold-change", ylab="-log P-adjusted", pch=20, cex=0.5)
points(resHTSeqShrunk$log2FoldChange[ highlight ], -log10(resHTSeqShrunk$pvalue[ highlight ]), col="green", pch=20, cex=0.5)
abline(v=0, h=-log10(0.01), lty="dashed", col="grey")
```

</details>

---

## Principal component analysis (PCA)

*DESeq2* provides a function to make a Principal Component Analysis (PCA) of the count data. The *DESeq2* [vignette](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-data-transformations) recommends using transformed counts as input to the PCA routines, as these transformations remove the dependence of the sample-to-sample variance on the genes' mean expression. 

One such transformations is the variance stabilizing transformation (VST). You can type `?varianceStabilizingTransformation` to learn more about this. To compare samples in an manner unbiased by prior information (i.e. the experimental condition), the `blind` argument is set to TRUE.

```{r}
transformed.vsd <- varianceStabilizingTransformation(ddsHTSeq, blind=TRUE)

plotPCA(transformed.vsd)
```

## Sample-to-sample correlation heatmap

Another common visualization of high-throughput datasets is a clustered heatmap of sample-to-sample distances (or correlations). This visualization groups togheter the samples that are more similar to each other. 

To make this visualization we first calculate a matrix of distances between all pairs of samples. Then we use the `heatmap` (from the base R package) to cluster and display the heatmap. 

```{r}
dists <- as.matrix(dist(t(normCounts)))
heatmap(dists, main="Clustering of sample-to-sample distances", scale="none")
```

We can also use pearson (or spearman) correlations as a distance metric. This is more robust than simple euclidean distances, and has the advantage that we can even use the raw (non-normalized) counts as input. It is generally a good idea to log transform the counts first.

```{r}
log10_rawCounts <- log10(counts(ddsHTSeq) + 1)
  
dists <- 1 - cor(log10_rawCounts, method="pearson")
heatmap(dists, main="Clustering of sample-to-sample pearson correlations", scale="none")
```

# Other visualizations

Here we plot the relative expression of all differentially expressed genes in the 6 samples. This figure is useful to visualize the differences in expression between samples. 

```{r, fig.height=8, fig.width=5}
library(gplots)

diffgenes <- rownames(resHTSeq)[ which(resHTSeq$padj < 0.05) ]
diffcounts <- normCounts[ diffgenes, ]

heatmap.2(diffcounts, 
          labRow = "", 
          trace = "none", density.info = "none",
          scale = "row",
          distfun = function(x) as.dist(1 - cor(t(x))))
```

The following commands are used to plot a heatmap of the 20 most differentially expressed genes. For this, we use the ordered results table to determine which genes are most differentially expressed, and then plot the values from the normalized counts table (transformed to log10).

```{r}
library(pheatmap)

# select the 20 most differentially expressed genes
select <- row.names(orderedRes[1:20, ])

# transform the counts to log10
log10_normCounts <- log10(normCounts + 1)

# get the values for the selected genes
values <- log10_normCounts[ select, ]

pheatmap(values,
         scale = "none", 
         cluster_rows = FALSE, 
         cluster_cols = FALSE,
         fontsize_row = 8,
         annotation_names_col = FALSE,
         gaps_col = c(3,6),
         display_numbers = TRUE,
         number_format = "%.2f",         
         height=12,
         width=6)
```

# Session information

```{r}
sessionInfo()
```

